---
title: "EDA"
---

# Introduction
=====================
This EDA is serves 4 purposes -
1. Sanity check on all variables to identify missing values and outliers.
2. Make transformations on variables if necessary
3. Clean up data set and rename the variables for consistency
4. Look for correlations in between variables to guide through the modeling stage

```{r results = "hide"}
library(dplyr)
library(ggplot2)
library(knitr)
library(corrplot)
```


```{r load data}
crime_data <- read.csv('../data/raw/crime_v2.csv')
```


# Looking at variables
========================
## 1. `county` is the county identifier, which is supposed to be one row per country identifier.

### Is there NA values?
```{r}
sum(is.na(crime_data$county))
crime_data[is.na(crime_data$county),]
sum(is.na(crime_data))
```
To me, all of them seems to be NA value for all variables. I will simply drop them as they don't provide any useful information.

```{r}
crime_data <- crime_data[is.na(crime_data$county) == FALSE, ]
```

### Look at the distribution of the values, are there outliers?

```{r}
table(crime_data$county)
```

However, looking at the count, I found that for county 193, there are 2 observations corresponding to it.

```{r}
crime_data[crime_data$county==193, ]
```
To me, it looks like pure duplication. So I will just drop one of the 2 observations.

```{r}
crime_data <- crime_data[-89, ]
crime_data[crime_data$county==193, ]
table(crime_data$county)
```

## 2. `year` is the year of the record.

### Look at the distribution of the values, are there outliers?
```{r}
table(crime_data$year)
```
It's worth it to transform it to a character variable as '1987'.
```{r}
crime_data$year <- '1987'
table(crime_data$year)
```


## 3. `crmrte` is the crimes committed per person
### Look at the summary of the numeric variable, is there any negative values that weren't supposed to be there?
```{r}
summary(crime_data$crmrte)
hist(crime_data$crmrte, breaks = 30)
```

Looking at the histogram, the crimes committed per person have a curve that skewed to right.

## 4. `prbarr` is the probability of arrest

### Look at the summary of the numeric variable, is there any negative values that weren't supposed to be there?
```{r}
summary(crime_data$prbarr)
```
Why is the max value greater than 1?
```{r}
crime_data[crime_data$prbarr>1,]
```
**Top coding** in this case. Since the probability should never exceeds 1 in theory, I will transform all the probability values that exceeds 1 into greater than 1.

```{r}
crime_data[crime_data$prbarr > 1,'prbarr'] <- 1
hist(crime_data$prbarr, breaks = 30)
```


## 5. `prbconv` is the probability of conviction

### Look at the summary of the numeric variable, is there any negative values that weren't supposed to be there?
```{r}
table(crime_data$prbconv)
# looks like `prbconv` is coded as a factor, I will change it to numeric
crime_data$prbconv <- as.numeric(levels(crime_data$prbconv))[crime_data$prbconv]
# check on the NA values
sum(is.na(crime_data$prbconv))
summary(crime_data$prbconv)
```
Same as above, why there are values greater than 1? Again, code them as 1 for the same reason
```{r}
crime_data[crime_data$prbconv > 1, 'prbconv'] <- 1
```


```{r}
hist(crime_data$prbconv, breaks = 30, main = '# of convictions/# of arrests')
hist(crime_data$prbconv*crime_data$prbarr, breaks = 30, main = '# of convictions/# of offenses')
```



## 6. `prbpris` is the probability of prison sentence

### Look at the summary of the numeric variable, is there any negative values that weren't supposed to be there?
```{r}
summary(crime_data$prbpris)
hist(crime_data$prbpris, main = '# of convictions resulting in prison/# of convictions', breaks = 30)
hist(crime_data$prbpris * crime_data$prbconv, main = '# of convictions resulting in prison/# of arrests', breaks = 30)
hist(crime_data$prbpris * crime_data$prbconv * crime_data$prbarr, main = '# of convictions resulting in prison/# of offenses', breaks = 30)
```
The histogram is very normally distributed

## 7. `avgsen` is the average sentence days

### Look at the summary of the numeric variable, is there any negative values that weren't supposed to be there?
```{r}
summary(crime_data$avgsen)
hist(crime_data$avgsen)
```
The minimum average sentence days ranging from 5 to 20 years

## 8. `polpc` is police per capita

### Look at the summary of the numeric variable, is there any negative values that weren't supposed to be there?
```{r}
summary(crime_data$polpc)
hist(crime_data$polpc, breaks = 30, main = 'Police per capita')
```

## 9. `density` is people per sq mile

### Look at the summary of the numeric variable, is there any negative values that weren't supposed to be there?
```{r}
summary(crime_data$density)
hist(crime_data$density, breaks = 30, main = 'People per sq mile by county')
hist(crime_data$density * crime_data$polpc, breaks = 30, main = 'Police per sq mile by county')
```

## 10. `taxpc` is tax revenue per capita

### Look at the summary of the numeric variable, is there any negative values that weren't supposed to be there?
```{r}
summary(crime_data$taxpc)
hist(crime_data$taxpc, breaks = 30, main = 'Tax Revenue per Capita')
```

## 11, 12, 13. `west` =1 if in western N.C, `central` =1 if in central N.C, `urban` =1 if in SMSA

### Look at the if it's mutually exclusive
```{r}
table(crime_data$west + crime_data$central)
table(crime_data$urban)
```
There is one observation that has 1 for both
```{r}
crime_data[crime_data$west + crime_data$central == 2, ]
```
I will just leave it there. One sample shouldn't have too much influence on the overall sample performance

## 14. `pctmin80` is the percentage of minority in 1980

### Look at the summary of the numeric variable, is there any negative values that weren't supposed to be there?
```{r}
summary(crime_data$pctmin80)
hist(crime_data$pctmin80, breaks = 30, main = 'Percentage of minority in 1980')
```

## 15-23. weekly wages
> `wcon` for construction, `wtuc` for transportation/utility/cummunication, `wtrd` for wholesale/retail, `wfir` for finance/investment/real estate, `wser` for service industry, `wmfg` for manufacturing, `wfed` for fed employees, `wsta` for state employees, `wloc` for local government employees

### Look at the summary of the numeric variable, is there any negative values that weren't supposed to be there?
```{r}
summary(crime_data[c('wcon', 'wtuc', 'wtrd', 'wfir', 'wser', 'wmfg', 'wfed', 'wsta', 'wloc')])
row_means <- apply(X=crime_data[c('wcon', 'wtuc', 'wtrd', 'wfir', 'wser', 'wmfg', 'wfed', 'wsta', 'wloc')], MARGIN=1, FUN=mean, na.rm=TRUE)
summary(row_means)
hist(row_means, breaks = 30, main = 'Average Weekly Wages across all files')
```
Here I noticed that `wser` has a maximum value of 2177.1, which is obviously an outlier, and seems to be recorded incorrectly.
```{r}
crime_data[crime_data$wser>2000, ]
```
It's safe to assume that, based on all other values, this value is supposed to be 217.7068 instead of 2177.068 so that it falls into the proper range of average weekly wage.
```{r}
crime_data[crime_data$wser>2000, 'wser'] <- crime_data[crime_data$wser>2000, 'wser'] / 10
summary(crime_data$wser)
```

#### create a new column containing the average wage across all sectors
```{r}
crime_data['avg_wage'] <- row_means
```


## 24. `mix` is offese mix: face-to-face/other

### Look at the summary of the numeric variable, is there any negative values that weren't supposed to be there?
```{r}
summary(crime_data$mix)
hist(crime_data$mix, main = 'Ratio of face-to-face offense vs other')
```

## 25. `pctymle` is percent young male

### Look at the summary of the numeric variable, is there any negative values that weren't supposed to be there?
```{r}
summary(crime_data$pctymle)
hist(crime_data$pctymle, breaks = 30, main = 'percentage of young male')
```




# Variable transformation
========================
Here, I'm adding 2 new variables: `pr_punishment` and `pr_conviction_over_offenses`, where
* `pr_punishment` represents the probability that offenses end up being sentenced to prison, or the ratio of convictions resulting in a prison sentence to offenses.
* `pr_conviction_over_offenses` represents the probability that offenses end up being convictions, or the ratio of convictions to offenses.
```{r}
crime_data['pr_punishment'] <- crime_data['prbarr'] * crime_data['prbconv'] *crime_data['prbpris']
crime_data['pr_conviction_over_offenses'] <- crime_data['prbarr'] * crime_data['prbconv']
```

```{r}
# renaming the variable names
colnames(crime_data) <- c('county_id', 'year', 'crime_rate', 'pr_arrest', 'pr_conviction', 'pr_sentence', 'avg_sentence',
                          'police_per_capita','population_density', 'tax_per_capita', 'is_west', 'is_central', 'is_urban',
                          'pct_minority','w_construction', 'w_tuc', 'w_sales', 'w_fin', 'w_service', 'w_manu', 'w_fed', 'w_state',
                          'w_local_gov', 'offense_mix', 'pct_young_male','avg_wage', 'pr_punishment', 'pr_conviction_over_offenses')
```

# Correlation Plot
========================
```{r}
crime_cols <- crime_data[c('crime_rate', 'pr_arrest', 'pr_conviction', 'pr_sentence', 'pr_punishment', 'pr_conviction_over_offenses',
                           'avg_sentence', 'police_per_capita',
                           'population_density', 'tax_per_capita', 'is_west', 'is_central', 'is_urban', 'pct_minority',
                           'w_construction', 'w_tuc', 'w_sales', 'w_fin', 'w_service', 'w_manu', 'w_fed', 'w_state', 'w_local_gov',
                           'offense_mix', 'pct_young_male', 'avg_wage')]

crime_cols_cor = cor(crime_cols)
corrplot(crime_cols_cor, type = "upper", tl.col = "black", tl.srt = 45, tl.cex = 0.7)
```
